Simple repression draft

### 2020-06-01 doc/section_01_introduction.tex
MRM pass through entire main text. Removed side notes. Still lots left

Gene expression presides over much of the most important dynamism of living
organisms.   The level of expression of batteries of different genes is altered
as a result of spatiotemporal cues that integrate chemical, mechanical and other
types of signals.  The original repressor-operator model conceived by Jacob and
Monod in the context of bacterial metabolism has now been transformed into the
much broader subject of gene regulatory networks in living organisms of all
kinds~\cite{Jacob1961, Britten1969, Ben-TabouDe-Leon2007}.  One of the remaining
outstanding challenges to have emerged in the genomic era is our continued
inability to predict the regulatory consequences of different regulatory
architectures. This challenge stems first and foremost from our ignorance about
what those architectures even are with more than 60\% of the genes even in an
ostensibly well understood organism such as  {\it E. coli} with no regulatory
insights at all~\cite{Rydenfelt2014-2,Belliveau2018,Ghatak2019,
Santos_Zavaleta2019}. But even once we have established the identity of key
transcription factors and their binding sites of a given promoter architecture,
there remains the predictive challenge of understanding its input-output
properties, an objective that can be met by a myriad of approaches using the
tools of statistical physics~\cite{Ackers1982, Shea1985,
Buchler2003,Vilar2003a,Vilar2003b, Bintu2005a,Bintu2005c, Gertz2009,Sherman2012,
Saiz2013, Ko1991,Peccoud1995,Record1996, Kepler2001,
Sanchez2008,Shahrezaei2008,Sanchez2011, Michel2010}.  One root to such
predictive understanding is to focus on the simplest regulatory architecture and
to push the theory-experiment dialogue as far and as hard as it can be
pushed~\cite{Phillips2019}.  If we demonstrate that we can pass that test by
successfully predicting both the means and variance in gene expression, then
that provides a more solid foundation upon which to launch into more complex
problems.

>>>>

Gene expression presides over much of the most important dynamism of living
organisms.   The level of expression of batteries of different genes is altered
as a result of spatiotemporal cues that integrate chemical, mechanical and other
types of signals.  The original repressor-operator model conceived by Jacob and
Monod in the context of bacterial metabolism has now been transformed into the
much broader subject of gene regulatory networks in living organisms of all
kinds~\cite{Jacob1961, Britten1969, Ben-TabouDe-Leon2007}.  One of the remaining
outstanding challenges to have emerged in the genomic era is our continued
inability to predict the regulatory consequences of different regulatory
architectures \mrm{, i.e. the arangement and effect of binding sites for
transcription factors in the DNA}. This challenge stems first and foremost from
our ignorance about what those architectures even are with more than 60\% of the
genes even in an ostensibly well understood organism such as  {\it E. coli} with
no regulatory insights at all~\cite{Rydenfelt2014-2,Belliveau2018,Ghatak2019,
Santos_Zavaleta2019}. But even once we have established the identity of key
transcription factors and their binding sites of a given promoter architecture,
there remains the predictive challenge of understanding its input-output
properties, an objective that can be met by a myriad of approaches using the
tools of statistical physics~\cite{Ackers1982, Shea1985,
Buchler2003,Vilar2003a,Vilar2003b, Bintu2005a,Bintu2005c, Gertz2009,Sherman2012,
Saiz2013, Ko1991,Peccoud1995,Record1996, Kepler2001,
Sanchez2008,Shahrezaei2008,Sanchez2011, Michel2010}.  One root to such
predictive understanding is to focus on the simplest regulatory architecture and
to push the theory-experiment dialogue as far and as hard as it can be
pushed~\cite{Phillips2019}.  If we demonstrate that we can pass that test by
successfully predicting both the means and variance in gene expression, then
that provides a more solid foundation upon which to launch into more complex
problems.

### 2020-06-01 doc/section_02_means.tex
RP
MRM pass through entire main text. Removed side notes. Still lots left

% !TEX root = ./busty_transcription.tex
\section{Means}\label{section_02_means}

@@

Fold-change has proven a very convenient observable in past
work~\cite{Garcia2011a, Razo-Mejia2018, Chure2019}. Part of its utility in
dissecting transcriptional regulation is its ratiometric nature, which removes
many secondary effects that are present when making an absolute gene expression
measurement. Also, by measuring otherwise identical cells with and without a
transcription factor present, any biological noise common to both conditions can
be made to cancel away. In addition, the presence of different operator sites,
i.e. binding sites for the repressor, is known to change the basal expression
level in the absence of repressor~\mmnote{look up cite}\mrm{does reg-seq mention
anything like this in their SI? I definitely saw that on my own experiments, but
didn't mention it in the paper}. Computing fold-changes neatly cancels this
effect as well.

@@

The master equation governing this model is analogous
to~\eqrange{eq:2state_cme_matrices}{eq:2state_rep_cme} for model 1 above. The
main subtlety arises since transcription initiation accompanies a promoter state
change. This can be understood by analogy to $\matr{K}$. The off-diagonal and
diagonal elements of $\matr{K}$ correspond to transitions arriving at or
departing from, respectively, the promoter state of interest. If transcription
initiation is accompanied by promoter state changes, we must have separate
matrices for arriving and departing transcription events since the arriving and
departing transitions have different initial copy numbers of mRNA, unlike for
$\matr{K}$ where they are the same. The master equation for this model is
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R_A} \vect{p}_{m-1}(t) +
                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
\label{eq:3state_rep_cme}
\end{equation}
with the state vector and promoter transition matrix defined as
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\ p_{m,P} \end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
                        0 & k_P^+ & -k_P^- 
                \end{pmatrix},
\label{eq:3state_cme_matrices_pt1}
\end{equation}
and the initiation matrices given by
\begin{equation}
\matr{R_A} = \begin{pmatrix}
                0 & 0 & 0 \\ 
                0 & 0 & r \\ 
                0 & 0 & 0
                \end{pmatrix},\
\matr{R_D} = \begin{pmatrix}
                0 & 0 & 0 \\ 
                0 & 0 & 0 \\ 
                0 & 0 & r
                \end{pmatrix}.
\label{eq:3state_cme_matrices_pt2}
\end{equation}
The elements of $\vect{p}_m$ encode the probabilities of having $m$ mRNA present
along with the promoter having repressor bound ($R$), being empty ($E$), or
having RNAP bound ($P$), respectively. $\matr{R_A}$ describes probability flux
\textit{arriving} at the state $\vect{p}_m$ from a state with one fewer mRNA,
namely $\vect{p}_{m-1}$, and $\matr{R_D}$ describes probability flux
\textit{departing} from the state $\vect{p}_m$ for a state with one more mRNA,
namely $\vect{p}_{m+1}$. $\matr{K}$ is closely analogous to model 1.

@@

\subsubsection{Nonequilibrium Model Four - ``Active'' and ``Inactive'' States}
This model, shown as (4) in Figure~\ref{fig1:means_cartoons}(C) is at the core
of the theory in~\cite{Razo-Mejia2020}. At a glance the cartoon for this model
may appear very similar to model 2, and mathematically it is, but the
interpretation is rather different. In model 2, we interpreted the third state
literally as an RNAP-bound promoter and modeled initiation of a transcript as
triggering a promoter state change, making the hopefully uncontroversial
assumption that an RNAP can only make one transcript at a time. In contrast, in
the present model the promoter state does \textit{not} change when a transcript
is initiated. So we no longer interpret these states as literally RNAP bound and
unbound but instead as coarse-grained ``active'' and ``inactive'' states, the
details of which we leave unspecified for now. We will comment more on this
model below when we discuss Fano factors of models.

@@

Why does the master curve of~\fig{fig1:means_cartoons}(D) exist at all? This
brings to mind the deep questions posed in, e.g.,~\cite{Frank2013}
and~\cite{Frank2014a}, asking us to search for general ``symmetries'' of the
problem instead of specifying generative models of the known microscopics of a
system. In our case, the key feature seems to be the exclusive nature of
repressor and RNAP binding, which allows the parameter describing the repressor,
$\Delta F_R$, to cleanly separate from all other details of the promoter
architecture, which are encapsulated in $\rho$. In the graph-theoretic language
which was applied to similar problems in~\cite{Gunawardena2012}
and~\cite{Ahsendorf2014}, the repressor-bound state is a leaf of the graph
describing transitions between promoter states. Arbitrary nonequilibrium
behavior can occur on the rest of the promoter state space, but it may all be
swept up in the effective parameter $\rho$, to which the repressor makes no
contribution.

@@

We have already pointed out that the master curve of~\fig{fig1:means_cartoons}
is essentially a one-parameter model, the one parameter being $\Delta F_R +
\log\rho$. By now the reader may be alarmed as to how can we even determine
$\Delta F_R$ and $\rho$ independently of each other, never mind shedding a lens
on the internal structure of $\rho$ itself. A hint is provided by the weak
promoter approximation, invoked repeatedly in prior studies~\cite{Bintu2005c,
Garcia2011a, Razo-Mejia2018} of simple repression using the 3-state equilibrium
model in~\fig{fig1:means_cartoons}(B). In that picture, the weak promoter
approximation means $\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P) \ll 1$,
meaning therefore $\rho\approx1$.  This approximation can be well justified on
the basis of the number of RNAP per cell and the strength of binding of RNAP to
DNA at weak promoters. This is suggestive, but how can we be sure that $\rho$
is not, for instance, actually $10^2$ and that $\Delta F_R$ hides a compensatory
factor? A resolution is offered by an independent inference of $\rho$ in the
absence of repressors. This was done in~\cite{Razo-Mejia2020} by fitting
nonequilibrium model 4 in~\fig{fig1:means_cartoons}(C), with zero repressor
(looking ahead, this is equivalent to model 4 in~\fig{fig2:constit_cartoons}), to
FISH data from~\cite{Brewster2014}. This provided a determination of $k^+$ and
$k^-$, from which their ratio is estimated to be no more than a few $10^{-1}$
and possibly as small as $10^{-2}$.

@@

Moving forward, we have weak evidence supporting the interpretation of $\Delta
F_R$ as a physically real free energy~\cite{Landman2019} and other work casting
doubt~\cite{Hammar2014}. How might we resolve the confusion? If there is no
discriminatory power to test the theory and distinguish the various models with
measurements of fold-changes in means, how do we probe the theory? Clearly to
discriminate between the nonequilibrium models in~\fig{fig1:means_cartoons}, we
need to go beyond means to ask questions about kinetics, noise and even full
distributions of mRNA copy numbers over a population of cells. If the
``one-curve-to-rule-them-all'' is more than a mathematical tautology, then the
free energy of repressor binding inferred from fold-change measurements should
agree with repressor binding and unbinding rates. In other words, the
equilibrium and nonequilibrium definitions of $\Delta F_R$ must agree, meaning
\begin{equation}
\Delta F_R = \beta\Delta\varepsilon_R - \log(R/N_{NS})
        = - \log(k_R^+/k_R^-),
\label{eq:deltaFR_eq_noneq_equiv}
\end{equation}
must hold, where $\beta\Delta\varepsilon_R$ is inferred from the master curve
fit to fold-change measurements, and $k_R^+$ and $k_R^-$ are inferred in some
orthogonal manner. Single molecule measurements such as from~\cite{Hammar2014}
have directly observed these rates, and in the remainder of this work we explore
a complementary approach: inferring repressor binding and unbinding rates
$k_R^+$ and $k_R^-$ from FISH measurements of mRNA population distributions.
\marginpar{\it MM: Add a notation aside somewhere that all association rates are written as zeroth-order rates. In other words, the concentration of the molecule is hidden inside, so for instance, $k_R^+$ is proportional to repressor copy number.}

>>>>

% !TEX root = ./busty_transcription.tex
\section{Mean Gene Expression}\label{section_02_means}

@@

Fold-change has proven a very convenient observable in past
work~\cite{Garcia2011a, Brewster2014, Razo-Mejia2018, Chure2019}. Part of its
utility in dissecting transcriptional regulation is its ratiometric nature,
which removes many secondary effects that are present when making an absolute
gene expression measurement. Also, by measuring otherwise identical cells with
and without a transcription factor present, any biological noise common to both
conditions can be made to cancel away. In addition, the presence of different
operator sites, i.e. binding sites for the repressor, is known to change the
basal expression level in the absence of repressor~\mmnote{look up
cite}\mrm{does reg-seq mention anything like this in their SI? I definitely saw
that on my own experiments, but didn't mention it in the paper}. Computing
fold-changes neatly cancels this effect as well.

@@

The master equation governing this model is analogous
to~\eqrange{eq:2state_cme_matrices}{eq:2state_rep_cme} for model 1 above. The
main subtlety arises since transcription initiation accompanies a promoter state
change. This can be understood by analogy to $\matr{K}$. The off-diagonal and
diagonal elements of $\matr{K}$ correspond to transitions arriving at or
departing from, respectively, the promoter state of interest. If transcription
initiation is accompanied by promoter state changes, we must have separate
matrices for arriving and departing transcription events since the arriving and
departing transitions have different initial copy numbers of mRNA, unlike for
$\matr{K}$ where they are the same (see Appendix~\cite{sec:non_bursty}). The
master equation for this model is
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R_A} \vect{p}_{m-1}(t) +
                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
\label{eq:3state_rep_cme}
\end{equation}
with the state vector and promoter transition matrix defined as
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\ p_{m,P} \end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
                        0 & k_P^+ & -k_P^- 
                \end{pmatrix},
\label{eq:3state_cme_matrices_pt1}
\end{equation}
and the initiation matrices given by
\begin{equation}
\matr{R_A} = \begin{pmatrix}
                0 & 0 & 0 \\ 
                0 & 0 & r \\ 
                0 & 0 & 0
                \end{pmatrix},\
\matr{R_D} = \begin{pmatrix}
                0 & 0 & 0 \\ 
                0 & 0 & 0 \\ 
                0 & 0 & r
                \end{pmatrix}.
\label{eq:3state_cme_matrices_pt2}
\end{equation}
The elements of $\vect{p}_m$ encode the probabilities of having $m$ mRNA present
along with the promoter having repressor bound ($R$), being empty ($E$), or
having RNAP bound ($P$), respectively. $\matr{R_A}$ describes probability flux
\textit{arriving} at the state $\vect{p}_m$ from a state with one fewer mRNA,
namely $\vect{p}_{m-1}$, and $\matr{R_D}$ describes probability flux
\textit{departing} from the state $\vect{p}_m$ for a state with one more mRNA,
namely $\vect{p}_{m+1}$. $\matr{K}$ is closely analogous to model 1.

@@

\subsubsection{Nonequilibrium Model Four - ``Active'' and ``Inactive'' States}
Model 4 in Figure~\ref{fig1:means_cartoons}(C) is at the core of the theory
in~\cite{Razo-Mejia2020}. At a glance the cartoon for this model may appear very
similar to model 2, and mathematically it is, but the interpretation is rather
different. In model 2, we interpreted the third state literally as an RNAP-bound
promoter and modeled initiation of a transcript as triggering a promoter state
change, making the hopefully uncontroversial assumption that an RNAP can only
make one transcript at a time. In contrast, in the present model the promoter
state does \textit{not} change when a transcript is initiated. So we no longer
interpret these states as literally RNAP bound and unbound but instead as
coarse-grained ``active'' and ``inactive'' states, the details of which we leave
unspecified for now. We will comment more on this model below when we discuss
Fano factors of models.

@@

Why does the master curve of~\fig{fig1:means_cartoons}(D) exist at all?
\textcolor{blue}{This brings to mind the deep questions posed in,
e.g.,~\cite{Frank2013} and~\cite{Frank2014a}, asking us to search for general
``symmetries'' of the problem instead of specifying generative models of the
known microscopics of a system} \mrm{I don't understand this last sentence. Can
we make it more transparent?}. In our case, the key feature seems to be the
exclusive nature of repressor and RNAP binding, which allows the parameter
describing the repressor, $\Delta F_R$, to cleanly separate from all other
details of the promoter architecture, which are encapsulated in $\rho$. In the
graph-theoretic language which was applied to similar problems
in~\cite{Gunawardena2012} and~\cite{Ahsendorf2014}, the repressor-bound state is
a leaf of the graph describing transitions between promoter states. Arbitrary
nonequilibrium behavior can occur on the rest of the promoter state space, but
it may all be swept up in the effective parameter $\rho$, to which the repressor
makes no contribution.

@@

We have already pointed out that the master curve of~\fig{fig1:means_cartoons}
is essentially a one-parameter model, the one parameter being $\Delta F_R +
\log\rho$. By now the reader may be alarmed as to how can we even determine
$\Delta F_R$ and $\rho$ independently of each other, never mind shedding a lens
on the internal structure of $\rho$ itself. A hint is provided by the weak
promoter approximation, invoked repeatedly in prior studies~\cite{Bintu2005c,
Garcia2011a, Razo-Mejia2018} of simple repression using the 3-state equilibrium
model in~\fig{fig1:means_cartoons}(B). In that picture, the weak promoter
approximation means $\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P) \ll 1$,
meaning therefore $\rho\approx1$.  This approximation can be well justified on
the basis of the number of RNAP per cell and the strength of binding of RNAP to
DNA at weak promoters. This is suggestive, but how can we be sure that $\rho$ is
not, for instance, actually $10^2$ and that $\Delta F_R$ hides a compensatory
factor? A resolution is offered by an independent inference of $\rho$ in the
absence of repressors. This was done in~\cite{Razo-Mejia2020} by fitting
nonequilibrium model 4 in~\fig{fig1:means_cartoons}(C), with zero repressor
(looking ahead, this is equivalent to model 4 in~\fig{fig2:constit_cartoons}),
to single-cell mRNA counts data from~\cite{Brewster2014}. This provided a
determination of $k^+$ and $k^-$, from which their ratio is estimated to be no
more than a few $10^{-1}$ and possibly as small as $10^{-2}$.

@@

Moving forward, we have weak evidence supporting the interpretation of $\Delta
F_R$ as a physically real free energy~\cite{Landman2019} and other work casting
doubt~\cite{Hammar2014}. How might we resolve the confusion? If there is no
discriminatory power to test the theory and distinguish the various models with
measurements of fold-changes in means, how do we probe the theory? Clearly to
discriminate between the nonequilibrium models in~\fig{fig1:means_cartoons}, we
need to go beyond means to ask questions about kinetics, noise and even full
distributions of mRNA copy numbers over a population of cells. If the
``one-curve-to-rule-them-all'' is more than a mathematical tautology, then the
free energy of repressor binding inferred from fold-change measurements should
agree with repressor binding and unbinding rates. In other words, the
equilibrium and nonequilibrium definitions of $\Delta F_R$ must agree, meaning
\begin{equation}
\Delta F_R = \beta\Delta\varepsilon_R - \log(R/N_{NS})
        = - \log(k_R^+/k_R^-),
\label{eq:deltaFR_eq_noneq_equiv}
\end{equation}
must hold, where $\beta\Delta\varepsilon_R$ is inferred from the master curve
fit to fold-change measurements, and $k_R^+$ and $k_R^-$ are inferred in some
orthogonal manner. Single molecule measurements such as from~\cite{Hammar2014}
have directly observed these rates, and in the remainder of this work we explore
a complementary approach: inferring repressor binding and unbinding rates
$k_R^+$ and $k_R^-$ from single-molecule measurements of mRNA population
distributions.
\marginpar{\it MM: Add a notation aside somewhere that all association rates are written as zeroth-order rates. In other words, the concentration of the molecule is hidden inside, so for instance, $k_R^+$ is proportional to repressor copy number.}
